abxD01 Paper

### 2015-03-26 PAPER.Rmd
Add hyp tests to compare CFU for dilutions

***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** All but the ciprofloxacin treatment resulted in similarly high levels of *C. difficile*. Therefore, we hypothesized that titrating the dose of antibiotics given to the mice would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C. difficile*. We titrated the doses of cefoperazone, streptomycin, and vancomycin given to the mice (Supplemental Table 1). Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis (Figure 3). *C. difficile* levels were significantly different between all titration groups within the cefoperazone (p<XXX0.03XXX) and streptomycin treated mice (p<XXX0.01XXX). For each of the antibiotics we titrated, reduced dosages resulted in reduced colonization by *C. difficile*. By sequencing the 16S rRNA genes from feces collected on day 0 from these mice, we sought to validate the relationships we observed with the full dosage. Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose. Most notable was the increase in the relative abundance of OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*. The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice. The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance. The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice. Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance. The relative abundance of many of the populations associated with resistance (*Porphyromonadaceae*, *Lachnospiraceae*, *Anaeroplasma*) increased with decreased vancomycin dosage; however, with the exception of the *Lactobacillus* populations, they did not reach the levels observed in the untreated mice. In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the *Akkermansia* and *Enterobacteriaceae*. Together, these results suggest that individual populations are not sufficient to provide colonization resistance. Rather, resistance is likely a product of the overall composition of the community.

![**Figure 3.**](results/figures/Figure3/figure3.png)

>>>>

```{r titration}
cfu <- read.table(file="data/process/abxD1.counts")
titration <- cfu[cfu$abx %in% c("vanc", "cef", "strep"),]

vanc <- titration[titration$abx=="vanc",]
vanc.test <- pairwise.t.test(log(vanc$CFU+1), factor(vanc$dose), method="BH")

strep <- titration[titration$abx=="strep",]
strep.test <- pairwise.t.test(log(strep$CFU+1), factor(strep$dose), method="BH")

cef <- titration[titration$abx=="cef",]
cef.test <- pairwise.t.test(log(cef$CFU+1), factor(cef$dose), method="BH")

***Reduced perturbation of the microbiota increases colonization resistance.*** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that reducing the dose of antibiotics that the mice received would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C. difficile*. In addition to the previous treatments, we treated mice with cefoperazone (0.3 and 0.1 mg/mL), streptomycin (0.5 and 0.1 mg/mL), and vancomycin (0.3 and 0.1 mg/mL). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin) and Gram-positive (i.e.  vancomycin) bacteria. Vancomycin was also selected because although the community was quite different from untreated mice, we did not observe high levels of colonization and we anticipated that lower doses might result in a community structure that would result in colonization. As expected, colonization levels decreased significantly in all mice receiving reduced doses of cefoperazone (P<`r round(max(cef.test$p.value, na.rm=T), 2)`) and streptomycin (P<`r round(max(strep.test$p.value, na.rm=T), 2)`; Figure 3). Although the 0.625 mg/mL dose of vancomycin did not result in colonization, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`).

Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose. Most notable was the increase in the relative abundance of OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*. The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.

The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance. The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.

Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance. The relative abundance of many of the populations associated with resistance (*Porphyromonadaceae*, *Lachnospiraceae*, *Anaeroplasma*) increased with decreased vancomycin dosage; however, with the exception of the *Lactobacillus* populations, they did not reach the levels observed in the untreated mice. In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the *Akkermansia* and *Enterobacteriaceae*.

Together, these results suggest that individual populations are not sufficient to provide colonization resistance. Rather, resistance is likely a product of the overall composition of the community.
